---
title: "Draft"
subtitle: "STA 210 - Project"
author: "Bayes' Harem - Christina Wang, Kat Cottrell, David Goh, Ethan Song"
format: pdf
editor: source
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load-pkgs
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(knitr)
library(ggfortify)
library(GGally)
```

```{r load-data}
#| message: false
abortion_data_full <- read_csv(here::here("data/abortion-attitudes", "wvs-usa-abortion-attitudes-data.csv"))
```

## Introduction

Paste-updated-content-here

## Data description

The dataset observes "attitudes on the justifiability of abortion in the
United States across six waves of World Values Survey data" (README.md)
and some basic qualities of the respondants.

Observations include:

-   WVS country code

-   Generational wave (1982, 1990, 1995, 1999, 2006, or 2011)

-   Justifiability of abortion (1-10)

-   Age (17 to 96)

-   College graduate (1 for yes)

-   Female (1 for women) - Unemployed (1 = currently unemployed)

-   Ideology (1-10 for left-right)

-   Financial satisfaction (1-10 for least-most)

```{=html}
<!-- -->
```
-   WVS post-materialist index (-1 = materialist. 2 = mixed. 3 =
    post-materialist)

-   Child autonomy index (-2 to 2 for obedience and religious
    faith-determination and independence)

-   Trust (1 = believes most people can be trusted)

-   Importance of God (1-10)

-   Opinion of respect for authority (-1-1 for bad to good)

-   National pride (1 = very proud to be an American)

The data were collected as part of the World Values Survey, which is
administered every few years and collects information about people's
values and beliefs worldwide. The survey aims to get a nationally
representative sample of a minimum of 1200 for most countries, and the
data are collected via face-to-face interviews at the respondents'
homes. The data included in this set specifically include responses from
6 waves of the survey (administered over the period 1982-2011). The
responses included in this set are from people in the United States, and
it examines their attitudes towards abortion.

## Analysis approach

Paste-updated-content-here

```{r filter-data}
abortion_data <- abortion_data_full %>% 
  filter(year >= "1995") %>%
  select(-starts_with("wave"), -starts_with("wvsccode"))
```

Before we select our predictor variables, we create a visualization and
summary statistics for the response variable.

```{r visualization-summary}
#| warning: false
ggplot(abortion_data, aes(x = aj)) +
  geom_histogram(binwidth = 1, aes(y=..density..)) + 
  geom_density(color = "red") +
  labs(title = "Histogram of abortion attitudes")

summary(abortion_data$aj)
```

We can observe from this histogram that the distribution of the outcome variable is not a bell shape, and it is trimodal. This is likely because the question's phrasing is similar to a yes/no question, but respondents were asked to give their level of agreement on a scale of 1-10. This may result in our model not being a good fit for the data if we attempt a multiple linear regression model. We have two backup plans for this, if our MLR eventually has a poor performance. First, we can truncate this data into a categorical outcome variable such as (Agree, Disagree, Undecided), and conduct a binomial or multinomial logistic regression. Second, we can filter our population based on various characteristics (if we have good reason to do so) so that our outcome data follows a bell shape.

We now conduct exploratory data analysis (EDA) on each of the 15
predictor variables in the data set. The EDA for each variable comprises
a histogram and a simple linear regression (SLR) against the outcome.

```{r eda-year}
#| warning: false
#Year of survey
ggplot(abortion_data, aes(x = year)) +
  geom_histogram() + 
  labs(title = "Histogram of year")
ggplot(abortion_data, aes(group = year, y = aj)) +
  geom_boxplot() 
```

```{r eda-age}
#| warning: false
#Age
abortion_data <- abortion_data %>% 
  mutate(agegroup = case_when(age >= 90 ~ '9',
                                             age >= 80  & age <= 89 ~ '8',
                                             age >= 70  & age <= 79 ~ '7',
                                             age >= 60  & age <= 69 ~ '6',
                                             age >= 50  & age <= 59 ~ '5',
                                             age >= 40  & age <= 49 ~ '4',
                                             age >= 30  & age <= 39 ~ '3',
                                             age >= 20  & age <= 29 ~ '2',
                              age <= 19 ~ '1'))
abortion_data
ggplot(abortion_data, aes(x = age)) +
  geom_histogram(binwidth = 1) + 
  labs(title = "Histogram of age")
ggplot(abortion_data, aes(x = agegroup, group = agegroup, y = aj)) +
  geom_boxplot()
```

```{r eda-college}
#| warning: false
# College education
ggplot(abortion_data, aes(x = collegeed)) +
  geom_histogram() + 
  labs(title = "Histogram of College education")
ggplot(abortion_data, aes(x = collegeed, group = collegeed, y = aj)) +
  geom_boxplot()
```

```{r eda-female}
#| warning: false
# Female
ggplot(abortion_data, aes(x = female)) +
  geom_histogram() + 
  labs(title = "Histogram of female")
ggplot(abortion_data, aes(x = female,group = female, y = aj)) +
  geom_boxplot()
```

```{r eda-unemployed}
#| warning: false
# Unemployed
ggplot(abortion_data, aes(x = unemployed)) +
  geom_histogram() + 
  labs(title = "Histogram of unemployed")
ggplot(abortion_data, aes(x = unemployed, group = unemployed, y = aj)) +
  geom_boxplot()
```

```{r eda-ideology}
#| warning: false
# ideology
ggplot(abortion_data, aes(x = ideology)) +
  geom_histogram() + 
  labs(title = "Histogram of ideology")
ggplot(abortion_data, aes(x = ideology,group = ideology, y = aj)) +
  geom_boxplot()
```

```{r eda-financial}
#| warning: false
# Financial satisfaction
ggplot(abortion_data, aes(x = satisfinancial)) +
  geom_histogram() + 
  labs(title = "Histogram of Financial satisfaction")
ggplot(abortion_data, aes(x = satisfinancial,group = satisfinancial, y = aj)) +
  geom_boxplot()
```

```{r eda-materialism}
#| warning: false
# WVS post-materialist index
ggplot(abortion_data, aes(x = postma4)) +
  geom_histogram() + 
  labs(title = "Histogram of WVS post-materialist index")
ggplot(abortion_data, aes(x = postma4, group = postma4,y = aj)) +
  geom_boxplot()
```

```{r eda-cai}
#| warning: false
# Child autonomy index
ggplot(abortion_data, aes(x = cai)) +
  geom_histogram() + 
  labs(title = "Histogram of Child autonomy index")
ggplot(abortion_data, aes(x = cai, group = cai, y = aj)) +
  geom_boxplot()
```

```{r eda-trust}
#| warning: false
# Trust most people
ggplot(abortion_data, aes(x = trustmostpeople)) +
  geom_histogram() + 
  labs(title = "Histogram of whether respondent trusted most people")
ggplot(abortion_data, aes(x = trustmostpeople, group = trustmostpeople, y = aj)) +
  geom_boxplot()
```

```{r eda-god}
#| warning: false
# Importance of God
ggplot(abortion_data, aes(x = godimportant)) +
  geom_histogram() + 
  labs(title = "Histogram of how respondent saw God's importance")
ggplot(abortion_data, aes(x = godimportant, group = godimportant, y = aj)) +
  geom_boxplot()
```

```{r eda-respect}
#| warning: false
# Respect for authority
ggplot(abortion_data, aes(x = respectauthority)) +
  geom_histogram() + 
  labs(title = "Histogram of respect for authority")
ggplot(abortion_data, aes(x = respectauthority, group = respectauthority, y = aj)) +
  geom_boxplot()
```

```{r eda-pride}
#| warning: false
# National Pride
ggplot(abortion_data, aes(x = nationalpride)) +
  geom_histogram() + 
  labs(title = "Histogram of National pride")
ggplot(abortion_data, aes(x = nationalpride, group = nationalpride, y = aj)) +
  geom_boxplot()
```

We now test if any of the predictors are strongly correlated with each
other.

```{r pair1}
#| warning: false
#I am doing ggpairs on 13 predictor variables. 
#To ensure they fit into the screen, I will make 4 matrices with subsets

ggpairs(abortion_data, 
  columns = c("year", "age", "collegeed", "female", "unemployed",
              "ideology", "satisfinancial"),
  columnLabels = c("year", "age", "college", "female", 
                   "unemployed", "ideology", "fin satis.")) +
   theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(angle = 45, size = 10),
    strip.text.y = element_text(angle = 0, hjust = 0)
    )
```

```{r pair2}
#| warning: false
ggpairs(abortion_data, 
  columns = c("postma4", "cai", "trustmostpeople", "godimportant", 
              "respectauthority","nationalpride"),
  columnLabels = c("postmaterialism", "child autonomy", "trust", "god", 
              "respectauthority","nationalpride")) +
   theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(angle = 45, size = 10),
    strip.text.y = element_text(angle = 0, hjust = 0)
    )
```

```{r pair3}
#| warning: false
ggpairs(abortion_data, 
  columns = c("year", "age", "collegeed", "female", 
              "respectauthority","nationalpride"),
  columnLabels = c("year", "age", "collegeed", "female", 
              "respectauthority","nationalpride")) +
   theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(angle = 45, size = 10),
    strip.text.y = element_text(angle = 0, hjust = 0)
    )
```

```{r pair4}
#| warning: false
ggpairs(abortion_data, 
  columns = c("postma4", "cai", "trustmostpeople", "godimportant", 
               "ideology", "satisfinancial"),
  columnLabels = c("postmaterialism", "child autonomy", "trust", "god", 
               "ideology", "satisfinancial")) +
   theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(angle = 45, size = 10),
    strip.text.y = element_text(angle = 0, hjust = 0)
    )
```

From these correlation matrices, we can conclude that the highest
correlations (above 0.2) are those between 
`godimportant` and `cai` 0.370 
`godimportant` and `respectauthority` 0.244 
`godimportant` and `ideology` 0.243 
`godimportant` and `nationalpride` 0.216
`nationalpride` and `age` 0.206 
`nationalpride` and `respectauthority` 0.206 
`respectauthority` and `cai` 0.205

We want to choose predictor variables that are not strongly correlated with each other, are relatively balanced in our data set, and visually appear to have a significant relationship with the outcome. Now that we have calculated the correlation matrix and know that some variables may have multicollinearity, we will select variables that do not exhibit significant multicollinearity for our model. Moreover, we can run several multiple linear regression models with different combinations of predictors to see which model performs best and is least affected by multicollinearity.

## Data dictionary

The data dictionary can be found [here](ADD%20LINK!!!).
